The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-03 kalman_walk_paper.tex
moved comments to margin
edits.

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

@@

@@

  \plr{Additional ideas to consider adding:
\begin{itemize}
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    \item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
\end{itemize}}

\plr{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}

@@

\plr{say in words what $\kappa$ is before the equation.

@@

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value \plr{sentence fragment}}
    \end{figure}
 % \end{example}

@@

@@

  %%%%%%%%%%%%%%%%%%%%%%%%
  \section*{Intraspecific variation and genetic drift}

\plr{Below we need to separate variation in $A$ from variation in phenotype.}

%  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}

%First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment. In this circumstance we expect phenotype to be conserved throughout evolutionary time. As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures. These phenotypic variations should yield distinct signatures. Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$. Genetic drift registers as an increase in the intrapopulation variation in $h(t)$. 
% \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}

  At any given time, there will be a range of regulatory coefficients present in the population
  due to segregating genetic variants.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?
  To answer these questions, we need to know 
  the strength of stabilizing selection on the phenotype,
  and the amount (and structure) of heritable variation in the genotype.
  These are known at best only roughly \citep{felsenstein1988phylogenies},
  so we aim for order-of-magnitude estimates.

@@

  We have guessed that within a population, 
  the entries of $A$ vary by around 1\%, 
  at least for networks whose function is strongly constrained.
  Subsequent generations for the most part resample from this diversity,
  so in a population of effective size $N_e$, with variation in $A$ of order $\sigma$,
  the population mean binding strengths
  will move a distance of order $\sigma/\sqrt{N_e}$ per generation,
  simply because this is the the variance of the mean of a random sample \citep{lande_drift}.
  % (This could be taken as a definition of $N_e$.)
  Selection will tend to push this mean towards the optimal set of networks \citep{ou_process},
  but mean movement parallel to the optimal set (that leaves the phenotype invariant) is unconstrained
  (although perhaps complicated by recombination load \citep{recomb_load}).
    \plr{recombination load won't slow it down actually I don't think}
  The action of genetic drift is also strongly determined by covariance between 
  standing genetic variation in different regulatory coefficients -- known as the $G$ matrix \citep{G_matrix},
  covariance which may arise due to functional constraints and/or statistical linkage.
  There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
  Linkage will almost certainly lead to covariance
  if the variation is due to \textit{cis}-regulatory variants,
  in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
  across which a population may carry only a few common haplotypes.
  However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
  a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
  (leading to positive covariance between $A_{ik}$ and $A_{jk}$)
  is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
  (Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
  In principle, there may be substantially less variation away from the set of optimal coefficients 
  than there is along the set, due to the action of selection.
  If so, this might require substantial epistatic load 
  -- mortality or reduced fitness of a large proportion of new offspring due to recombination between somewhat incompatible alleles.
  However, it is unclear if this is likely to occur and XXX we revisit this below.

  It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
  (a) a ``cloud'' of width X about the population mean, 
  which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
  In fact, the population mean will not produce exactly the optimal phenotype,
  but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.

  %%%%%%%%%%%%%%%%%%%%%%%%%%%
  \paragraph{Brownian motion} on the set of equivalent networks:
    \plr{revisit.  should this describe motion \emph{on} the set, or motion \emph{near} it, as in writeup with todd?}
  The set $\calA$ is characterized as the solutions to the equations \eqref{eqn:equivalence},
  and is hence an algebraic variety. 
  In fact, the Kalman decomposition \plr{XXX above} provides us with an explicit characterization of the set.
  Above we argued that the population mean set of coefficients $A$ was subject to genetic drift, 
  with each entry $A_{ij}$ changing with mean square displacement $\sigma^2/\sqrt{N_e}$ per generation.
  Since selection constrains the population mean to stay near to the set of optimal networks $\calA$,
  to a good approximation, the population mean moves as Brownian motion in the space of matrices $\R^{n \times n}$
  but conditioned to stay on the optimal set.
  This Brownian motion on $\calA$ is a stochastic process driven by the Laplace-Beltrami operator on $\calA$
  where this makes sense, with special behavior at the singular points (if any \plr{XXX}).
  Unlike Brownian motion in flat space, this stochastic process can have \emph{bias} --
  for instance, it is pushed away from regions of negative intrinsic curvature
  (because there is ``less space'' in such regions).

  \plr{NOT TRUE?:}
  However, the process always has the property that
  changes in $A$ accumulate at constant (mean squared) rate:
  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
  \begin{align}
      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 T + O(T^2) .
  \end{align}

  %\begin{align}
  %  \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
  %  A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
  %\end{align}
 % \section*{Phenotypic Invariance}

%    \begin{example}[Not all minimal gene networks can drift]
 %     If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.

   %  \begin{align}
   %     C  = I \lor B = I \\
   %     IV^{-1} &= I \\
   %     \iff V &= I
   %   \end{align}
 %     \end{example}
  %    \begin{example}[All Non-Minimal Gene Networks can Drift]
  %    
  %    Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
  %    \begin{align}
  %      h(t) &= \widehat{h}(t) \iff \\ 
  %      CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
  %    \end{align}
  %    Any two systems with equivalent impulse responses will have equivalent phenotypes.
  %    \end{example}

      \subsection*{Speciation via Reproductive Incompatibility}

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
    Note the difference in scale on the $y$-axis. F2 hybrids display more phenotypic divergence than F1s, on average. Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
    \plr{make axis labels bigger}
  }
\end{figure}

@@

Roughly, what is this doing?
Suppose that the population mean differs from the optimum by $\epsilon$,
that $\Sigma = \sigma^2 I$ and $U = s I$.
Then the population mean gets closer to the optimum on average, moving to
$\epsilon/(1 + \sigma^2/s)$
and adds noise of size $\sqrt{s} \sigma/\sqrt{N \sigma^2 + N s}$.
At equilibrium, these two movements will be of the same order,
so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2/s}$.

>>>>

% for comments in margins or not
\newif\ifmargincomments
\margincommentstrue
% \margincommentstrue
\ifmargincomments
    \usepackage{todonotes}
    \usepackage[left=1cm,right=6.5cm,top=3cm,bottom=3cm,nohead,nofoot,marginparwidth=6cm]{geometry}
    \newcommand{\plr}[1]{\todo[color=blue!25]{#1}}
    % inline comment version
    \newcommand{\plri}[1]{{\color{blue}\it #1}}
\else
    \newcommand{\plr}[1]{{\color{blue}\it #1}}
    \newcommand{\plri}[1]{\plr{#1}}
\fi

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

@@

% fitness as a fn of distance
\newcommand{\fit}{\mathcal{F}}
% fitness as a fn of A
\newcommand{\fitx}{\mathcal{f}}
% set of optimal coefficients
\newcommand{\optx}{\mathcal{X}}
% optimal phenotype
\newcommand{\optph}{\Phi_0}
% distance in phenotype space
\newcommand{\dph}{d}

@@

\plr{Additional ideas to consider adding:}
\begin{itemize}
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    \item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
\end{itemize}

\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}

\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}

@@

\plr{say in words what $\kappa$ is before the equation.}

@@

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. 
        $\tau$ can take any value \plri{sentence fragment}}
    \end{figure}
 % \end{example}

@@

\plri{Missing: a statement about when there are neutral directions, and how many.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}

At any given time, there will be a range of regulatory coefficients present in the population
due to segregating genetic variants.
Over many generations, even if selective pressures do not change,
this range of networks will shift 
as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
How much variation do we expect to find within a population?
Is this range limited by available variation or kept in check by selection?
How fast will a population explore the space of equivalent networks?
In this section we explore informally a general model for this situation,
in which a population drifts stochastically near a set of equivalent, optimal systems.
We work with a population of effective size $N_e$.

Since the argument applies more generally, we abstract the model above.
Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \fit(\dph(\Phi(x),\optph)/\beta)$,
where $\beta$ controls the scale on which fitness drops.
We will assume that the map $\Phi$ and the distance function $\dph()$ are smooth,
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

\paragraph{Offspring.}
Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
and the individual's coefficients are the average of her two haploid values (no dominance).
This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
Each new gamete is produced from the parent's two haploid copies;
for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
and so $\sigma$ remains constant, up to a $1/N_e$ term.
A more general model including segregation variance \citep{barton_infinitesimal}
would result in the same conclusions.
\plri{but put this in the appendix?}

\paragraph{System drift}
% plr how fast does it drift
If the variation within a population of some coefficient
has standard deviation $\sigma$,
then since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but mean movement along the optimal set $\optx$ is unconstrained.
% (although perhaps complicated by recombination load \citep{recomb_load})
The value of $\sigma$ in particular directions in coefficient space 
depends on constraints imposed by selection and 
the covariance between genetic variation between different coefficients -- known as the $G$ matrix \citep{G_matrix}.
% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
For instance, linkage will lead to covariance
if the variation is due to \textit{cis}-regulatory variants,
in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
across which a population may carry only a few common haplotypes.
However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

% It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
% (a) a ``cloud'' of width X about the population mean, 
% which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
% In fact, the population mean will not produce exactly the optimal phenotype,
% but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.

To obtain a general quantitative picture, we need to know 
$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively;
$\gamma$, how fast phenotype changes moving away from $\optx$;
and $\beta$, how fast fitness drops with distance from the optimal phenotype.
(Concretely, $\gamma$ is the derivative of $\dph(\Phi(x+u z), \Phi(x))$ with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent subspace at $x$.)
With these parameters, a typical individual will have a fitness of around $\fit(\sigma_S \gamma / \beta)$.

\paragraph{Hybridization}
\plr{I put F1s and F2s here, but with less explanation than below. Merge somehow.}
By the arguments above, 
over $T$ generations the population mean will move a distance of order $\sqrt{\sigma_N/T}$ along $\optx$.
A $F_1$ hybrid between the two populations
has one haploid genome from each,
so will be distributed as the mean of two randomly chosen haploids from the two populations,
and so will have mean at the midpoint between their means,
and variance equal to $\sigma$.
A $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations.
These are depicted in figure \ref{fig:conceptual_fig}.

\begin{figure}
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    WRITE CAPTION
    \plri{The figure is not quite right now -- should show all types of F2 in purple.  but then it will be messy.  ideas?}
}
\end{figure}

\plri{improve figure by putting labels on from the following}
\plr{I think we can set $\beta=1$ by defining $\fit$ appropriately?}
Suppose that two populations have drifted independently to differ by $z$,
and that $z$ is of the same order as $\sigma$ but is smaller than $\gamma$,
so that for $x_0$ a point nearby on $\optx$,
\begin{align*}
    \Phi(x_0 + x) = \Phi(x_0) + D \Phi(x_0) \cdot x + O(\|\gamma x\|^2),
\end{align*}
and so $d(\Phi(x_0+x), \Phi(x_0)) \approx  \gamma x$.
The mean $F_1$ is the average of the parental means,
and since the first-order terms in the Taylor series vanish,
has phenotype differing from the optimum by a distance of order $(\gamma z)^2$.
\plri{(Proof: let the parental means be $x_1$ and $x_2$; then as $z=x_2-x_1$,
    $d(\Phi((x_1+x_2)/2), \Phi(x_1)) = D \Phi(x_1) \cdot z/2 + O(\gamma z)^2$,
    but expanding about $x_2$ instead we get
    $d(\Phi((x_1+x_2)/2), \Phi(x_2)) = - D \Phi(x_2) \cdot z/2 + O(\gamma \|z\|)^2$.
    Since $\Phi(x_1) = \Phi(x_2)$ and $D \Phi(x_1) - D \Phi(x_2) = O(\|z\|^2)$,
    we see that $D \Phi(x_1) \cdot z = O(\gamma z)^2$.}
The mean $F_2$ is the same,
but is a distance away from the optimal set of order $z$.
In summary, the typical fitness of an individual in the original population is
$\fit_0 = \fit(\sigma_s \gamma / \beta)$;
of an $F_1$ is
$\fit_1 = \fit((\sigma_s \gamma + \gamma z^2) / \beta)$;
and of an $F_2$ is
$\fit_2 = \fit(\gamma z / \beta)$.

If $\fit(x) = e^{-x^2} \approx 1 - x^2$,
then ...
\plri{reparameterize to define $\fit_1$ as the \emph{drop} in fitness?}

%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Parameter estimates for real systems}

To translate these results into real predictions, we need to know
the strength of stabilizing selection on the phenotype,
and the amount (and structure) of heritable variation in the genotype.
These are known at best only roughly \citep{felsenstein1988phylogenies},
so we aim for order-of-magnitude estimates.

@@

We have guessed that within a population, 
the entries of $A$ vary by around 1\%, 
at least for networks whose function is strongly constrained.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Speciation via Reproductive Incompatibility}

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
    Note the difference in scale on the $y$-axis. F2 hybrids display more phenotypic divergence than F1s, on average. Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
    \plri{make axis labels bigger}
  }
\end{figure}

@@

Roughly, what is this doing?
Suppose that the population mean differs from the optimum by $\epsilon$,
that $\Sigma = \sigma^2 I$ and $U = I/\beta^2$ (so, stablizing selection happening on a distance scale of $\beta$).
Then the population mean gets closer to the optimum on average, moving to
$\epsilon/(1 + \sigma^2 \beta^2)$
and adds noise of size $(1/\beta) \sigma/\sqrt{N \sigma^2 + N 1/\beta^2}$.
At equilibrium, these two movements will be of the same order,
so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2 \beta^2}$.
